{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-05-29 Docs/manuscript.tex
updated intro

\newpage

\section*{Abstract}

% We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.

\section{Introduction}
To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}. One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}. Local adaptation - how well the growth and adaptive characteristics of plants align with the local climate conditions - influences their ability to thrive and reproduce  \citep{casmey22, kawecki04, anderson12, kim13}. When the selection intensity exceeds gene flow in species that spread over diverse areas, clinal variations among environmental factors can result in populations with unique adaptations to the local climate and, therefore, greater fitness \citep{kremer12, rossi15, salmela13}. On the other hand, phenotypic plasticity describes an individual genotype’s ability to exhibit different phenotypes in response to environmental variations, enabling plants to adjust to new environmental conditions and survive in this rapidly changing world (also known as ‘plastic rescue’)  \citep{chevin10, chevin102, snell18}. \\
\\
Among the various population-level characteristics critical in synchronizing plant growth with the available growing season, phenology traits hold particular significance \citep{primack80, oneil97, chuine01}. Phenology governs the timing of plant transitions between dormancy and active growing, allowing them to time reproduction and utilize each growing season as best as possible \citep{chuine10, hanninen11}. Additionally, phenology provides insight into how climate change affects plant life cycles, regeneration capacity, and ecosystem management.\\
\\
Spring phenology, specifically the timing of budburst and leaf flush, plays a significant role in determining plant fitness and has been widely adopted as a useful tool to investigate plant adaptation mechanisms \citep{guo22, chuine01}. In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}. Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}. Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}. \\
\\
In recent decades, there has been growing interest in how climate change influences plant phenology \citep{Gill15}. Trees exhibit immediate responses via phenotypic plasticity with the warming climate \citep{Nicotra10}, while the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. Phenological shifts in spring phenology due to recent temperature increases have been extensively observed \citep{Parmesan06}. For example, leaf unfolding in Europe advanced up to six days since the early 1960s due to the rise in spring temperature, and spring events have generally shifted more than fall events \citep{menzel99}. Studies have found that species that shift more with warming also appear to perform better \citep{Gill15}. For instance, research has demonstrated that the timing of budburst and senescence in deciduous plants can impact plant competition \citep{fridley12}, plant growth \citep{myneni97}, and the uptake of carbon by the ecosystem \citep{Barichivich12}. As leaf phenology shifts in response to climate, increasing evidence suggests how much a species shifts its phenology with warming may predict fitness consequences: Individuals, populations, and phenological events (spring versus fall) with more plasticity might fare better with climate change. Plasticity, however, likely varies across species and phenophases. An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction. \\
\\
Much spring phenology is driven by plasticity. For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through common garden experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}. Over time, comparable clinal variations have been observed in these experiments along latitudinal and elevational gradients for many temperate and boreal species. Fall bud set phenology and cold hardiness exhibit strong population differentiation and strong clines along different latitudes, altitudes, and continentality gradients \citep{Alberto13}. For example, low-latitude populations have shown significantly stronger delays in leaf senescence over time than high-latitude populations \citep{Gill15}. Spring events (budburst, leaf flush, etc.) show more substantial phenotypic plasticity and higher variability \citep{AitkenBemmels16}.\\
\\
Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies. We ask: To what extent are patterns of phenotypic plasticity and local adaptation along climate gradients similar across common garden studies? To better understand these clines, we also aim to examine the following:
\begin{itemize}
\item How different are clines between spring and fall phenology? 
\item How different are clines across Europe and North America?
\item If differences in clines exist across angiosperm and gymnosperm species?
\item If climate overlap between the provenance and garden explains the similarities and differences in clines?
\end{itemize}
To address these questions, we reanalyzed data from the literature on common garden experiments in temperate tree species across Europe and North America.

\section{Methods}
\subsection{Data Collection}
To locate common garden studies that reported the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature. On 14 December 2022 we searched Web of Science (Thompson Reuters, New York, NY) using the following terms:
\begin{quote}
TOPIC = (common garden* OR provenance*) AND (leafout* OR leaf out* OR budburst OR spring phenolog*)
\end{quote}
which returned 122 publications. We also contacted authors of previous review papers \citep{AitkenBemmels16, Alberto13}, to help further search the literature. We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported latitude and longitude of provenances and gardens, and (d) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see \nameref{supplement-section:addmethods} in the Supplements).
\\
\\
Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the 'lubridate' package in R \citep{Grolemund11}.
\\
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (number of studies = 12). See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.} 
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}. To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the 'daymetr' package in R respectively \citep{cornes2018,hufkens2018}. Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the 'overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020. Finally, using the daily temperature data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).

@@

\section{Results}
Our final dataset included seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species also had fall event information available. 
\\

Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig. \ref{figure:springfall_latmat}, Table. \ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor in North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT:-0.16 days/$^{\circ}$C [-0.55 - 0.23]). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden). \\
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events WEre earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continental_spptype_effect}A). \\

@@

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
Spring events slightly delayed 0.37 (0.15 - 0.59) days per degree north in angiosperms and 0.23 (0.00 - 0.46) days per degree north in gymnosperms. Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
However, effects of MAT on spring events weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig. \ref{figure:continental_spptype_effect}B).\\

>>>>

\newpage
% Journal ideas: Global Change Biology; Functional Ecology or Journal of Ecology next seem good ... maybe a biogeography journal (though I am not sure if we can afford one) ... maybe New Phytologist but I think they could be picky.

% Just jotting down and putting some bits you have for abstract for now. 
\begin{abstract}
Climate change forecasting .... plasticity.... adaptation. Spring tree phenology is critical to forecasting ... Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events. We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.  We found ... This is important because... 
\end{abstract}

\section{Introduction}
%emw: FYI this is a first draft -- meaning it still has lots of words, sentences and other bits to fix; but best to do that as we go. 
%emw: I noticed a lot of papers on local adaptation don't always use the term 'local adaptation' so we may want to avoid using it too much? I suspect it may be contested. I also wonder if my advice about definitions was wrong ... but let's definitely save that text for now. 
%emw: We need to 'funnel' in our intros -- until we submit to a plant journal focusing on plants too early is getting too narrow too fast.

% To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}. One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}. 
The increasing need to predict the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}. In particular, the relative importance of genetic adaptation versus plasticity is of paramount importance for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptations \citep{chevin102, snell18}.

% Local adaptation---how well the growth and adaptive characteristics align with the local climate conditions---influences their ability to thrive and reproduce  \citep{casmey22, kawecki04, anderson12, kim13}. When the selection intensity exceeds gene flow in species that spread over diverse areas, clinal variations among environmental factors can result in populations with unique adaptations to the local climate and, therefore, greater fitness \citep{kremer12, rossi15, salmela13}. On the other hand, phenotypic plasticity describes an individual genotype’s ability to exhibit different phenotypes in response to environmental variations, enabling plants to adjust to new environmental conditions and survive in this rapidly changing world (also known as `plastic rescue')  \citep{chevin10, chevin102, snell18}.

Many of the currently observed responses to climate change appear to be mostly plastic \citep{burton22,zettle21,bonamour19, king17}, including the most reported biological response to climate change---shifting phenology. Phenology (timing of recurring seasonal events) governs the timing of transitions between dormancy and active growth for many organisms, allowing them to time reproduction and fully exploit the resources of each growing season \citep{chuine10,hanninen11}. Many terrestrial animals time their breeding season to coincide with the availability of food resources. For example, migratory birds arrive at breeding grounds amidst the peaking of insects \citep{hallfors20}. Many butterfly species ensure their larvae have an ample food supply by synchronizing their egg-laying with the phenology of their host plants \citep{rytteri21,posle18}. Similarly, the migration and spawning of many marine organisms are closely tied to water temperature and the availability of planktonic food sources \citep{genner10}. Such synchronization enables them to provide optimal food for their young during critical growth stages. Therefore, climate warming and the associated changes in spring phenology timing, such as earlier leafout or insect emergence due to climate change, can disrupt this synchronization and negatively impact animal populations \citep{wann19,renner18, chu17}. As such, phenology plays a significant role in determining fitness \citep{guo22, chuine01}. Shifted phenology in recent decades---with many events moving several days per decade \citep{vita21,khar18,Menzel06} ---has thus led to concerns about fitness consequences, and the limits of possible future shifts. Future phenological shifts will depend on how much phenology is determined by adaptation versus plasticity will determine, yet our understanding of the balance of these two approaches to variation is limited, even for species groups that are critical to both forecasting and have been well studied, such as trees.

Tree phenology is critically important to climate change forecasting and has been an area of long-term study of adaptation and plasticity \citep{Cleland:2007or}. The timing of budburst and senescence can impact plant competition \citep{fridley12}, growth \citep{myneni97}, and carbon sequestration \citep{Barichivich12}---thus affecting climate change itself. Hundreds of years of study have established common clines in fall phenology, as locations with shorter growing season (poleward, higher elevation) exhibit earlier stopping of growth (e.g., via budset), and connected these to an underlying proximate mechanism of changing photoperiod cues, driven by local adaptation to the local growing season and relevant photoperiod to stop growth \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows important levels of adaptation.

Evidence for local adaptation in spring phenology comes mostly from common garden studies. For more than 250 years, researchers have been comparing populations of trees from different geographical origins under the same environmental conditions through these experiments to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}. Studies have documented provenance differences of 2.08-4.17 days per degree latitude for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (For example,\emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to some debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they appear potentially more common in European studies.

Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}. Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years. Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{cat23, ospreebbms,ospreephoto}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}. Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.  
To test for evidence of adaptation in spring phenology and what factors may underlie differences observed across studies, we comprehensively examined clines for spring and fall events. We test for evidence of adaptation via provenance trends with latitude and climate. Further, to better understand these clines, we also examine the following:
\begin{itemize}
\item How different are clines between spring and fall phenology? 
\item How different are clines between Europe and North America?
\item How different are clines between angiosperm and gymnosperm species?
\end{itemize}
To address these questions, we combined a new meta-analysis of all common garden experiments in temperate tree species across Europe and North America reporting spring phenology with Bayesian hierarchical models.

 %REFS
 % savolainenAREES -- have not read recently but sounds relevant
 % gauzere2020 -- cite and read
 % \cite{chang2021} and \cite{chuine2016} -- relevant for discussion of chilling
 % See also Salk 2020 paper and some papers mentioned above

\section{Methods}
\subsection{Data Collection}
To locate common garden studies that reported the timing of spring events of woody plant species we searched and reviewed the peer-reviewed literature. On 14 December 2022 we searched Web of Science (Thompson Reuters, New York, NY) using the following terms:
\begin{quote}
TOPIC = (common garden* OR provenance*) AND (leafout* OR leaf out* OR budburst OR spring phenolog*)
\end{quote}
which returned 122 publications. We also contacted authors of previous review papers \citep{AitkenBemmels16, Alberto13}, to help further search the literature. We then reviewed the methods and results of all publications to refine to only studies that met the following criteria: (a) focused on woody plants originating from either Europe or North America, (b) had provenance trials/common gardens on the same continent, (c) reported latitude and longitude of provenances and gardens, and (d) reported spring events in units of calendar days (day of year or DOY) or could be converted into DOY (see \nameref{supplement-section:addmethods} in the Supplements).

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the 'lubridate' package in R \citep{Grolemund11}.
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (number of studies = 12). See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.} 
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}. To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the 'daymetr' package in R respectively \citep{cornes2018,hufkens2018}. Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the 'overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020. Finally, using the daily temperature data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).

@@

\section{Results}
Our final dataset included seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species also had fall event information available.

Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig. \ref{figure:springfall_latmat}, Table. \ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor in North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT:-0.16 days/$^{\circ}$C [-0.55 - 0.23]). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

In contrast, fall events (budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events WEre earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continental_spptype_effect}A).

@@

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
Spring events slightly delayed 0.37 (0.15 - 0.59) days per degree north in angiosperms and 0.23 (0.00 - 0.46) days per degree north in gymnosperms. Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
However, effects of MAT on spring events weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig. \ref{figure:continental_spptype_effect}B).
